Specification of Advanced Types Translation

### 2014-03-05 DOCUMENTATION/AdvancedTypes.tex
Doing the first review by me.

\author{David Déharbe, Valério Medeiros Jr.}
\title{Array}

@@

The supported advanced types are array and structures. The nodes Array and 
Structure can be used to declare a variable or to select an element in the array 
or the structure.

@@

    {\footnotesize

@@

        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Record}}\\
	\multicolumn{3}{c}{\B{Vari}\alt\B{StructItem}\alt\B{ArrayItem} } \\

@@

        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\

@@

\begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
        \hline
	\B{base} & : &  \B{Struct} \\
	\B{element} & : & \B{Name}\\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record Structure : \B{Record}} \\
        \hline
	\B{elements} & : & seq \B{Name}\\
	\B{values} & : & seq \B{Expr}\\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

\VGM{To add in text: The nodes \B{Int\_lit} and \B{Bool\_lit} are literals in its 
respective set.}
\VGM{The LLVM grammar with support to array multidimensional.}

@@

The rule lvalue is:
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an array item} \THEN  \trad{n}{ArrayItem} \\
& \quad \IF \mbox{\B{n} is a struct item} \THEN  \trad{n}{StructItem} \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

@@

Two approaches are possible to support arrays and structures. An approach is importing reusable libraries of AtelierB that are reported in the official document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. This approach is supported, but the generated code is more complex and it needs more intermediate steps to manipulate the data of imported components. Moreover, this approach allows dynamic allocation for arrays, but it is strongly not recommended to safety-critical systems. A second approach enjoys the direct correspondence between B0 native arrays and LLVM-IR and this approach is presented here.

\subsection{Structure}
The struct declaration defines a list of variables to be placed under one name, 
allowing several variables to be accessed by one base name. This example get andset one value of xx'balance and initializes a structure.

\begin{ccode}
xx := rec(name:aa, balance:10) ;
xx'balance := xx'balance -1;
\end{ccode}

The following rule get an element value from the struct. The same rule is used in left side of attribution as a $LValue$ and right side of attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.

\begin{align*}
\begin{split}
  \trad{n}{\B{StructItem}} \isdef \\
   & \LET \ell = \lbl \AND \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   & \llvm{"\PH{\ell} = getelementptr \%\PH{t} @\PH{v}, i32 0, i32  \SBPos(\B{n})"} \\
   & \sep  \ell  \sep \SBType(\B{n})  \\
\end{split}
\end{align*}

The attribution from record structure to a structure variable, line 1 from last code example, is translated by a sequence of application of rule $StructItem$.

\section{Arrays}
A concrete array is a total function where the domain indexes the range 
elements. The domain is composed of Cartesian product of sequence of concrete 
sets. The node Array in AST represents a declaration of array or a 
selection of element.

Some examples of array are:
\begin{itemize}
	\item $array1 \in (0..10) \to INT $	(a simple array)
	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
	\item $\{1\vdash1\vdash4, 1\vdash2\vdash5\}  \in (INT \times INT) \times INT$ (an initialised array)
\end{itemize}

\subsection{Relation arrays B0 and LLVM-IR}

@@

This section shows the relation between arrays B0 and LLVM-IR. The sequence of
steps are: defining  an array, getting an element and setting an
element.

@@

The essential elements of the array are its name, size and type. 
The size of ``e'' array with the indexes (i1,...,in) are defined according 
to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
e.in\_start+1)$.

@@

\VGM{It is not implemented yet, because it requires generating an XML example.}
The following example initialises a new array in C:

@@

The syntax rule used to initialize an array in LLVM is:\\
$@<var\_name> = constant [<size> x <type>] [<type> <value>]$.\\
The translation of the last example is:

@@

The LLVM-IR instruction \textit{getelementptr} is used to select an element in
the array or structure. A simple example is:

@@

The following blocks illustrate a C code setting a value to array
and its translation to LLVM-IR.

@@

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.

\begin{llvmcode}
%x = alloca i32, align 4
%y = alloca i32, align 4
%z = alloca i32, align 4
store i32 2, i32* %x, align 4
store i32 3, i32* %y, align 4
store i32 4, i32* %z, align 4
%1 = load i32* %z, align 4
%2 = sext i32 %1 to i64
%3 = load i32* %y, align 4
%4 = sext i32 %3 to i64
%5 = load i32* %x, align 4
%6 = sext i32 %5 to i64
%7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7, align 4;setting an element
\end{llvmcode}

@@

\subsection{A  simple example and its LLVM-IR}

@@

define void @Array__set(i32 %ix, i32 %tt) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32, align 4
  store i32 %ix, i32* %1, align 4
  store i32 %tt, i32* %2, align 4
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %19

; <label>:5                                       ; preds = %0
  %6 = load i32* %1, align 4
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %19

; <label>:8                                       ; preds = %5
  %9 = load i32* %2, align 4
  %10 = icmp sge i32 %9, 0
  br i1 %10, label %11, label %19

; <label>:11                                      ; preds = %8
  %12 = load i32* %2, align 4
  %13 = icmp sle i32 %12, 1000
  br i1 %13, label %14, label %19

; <label>:14                                      ; preds = %11
  %15 = load i32* %2, align 4
  %16 = load i32* %1, align 4
  %17 = sext i32 %16 to i64
  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
  store i32 %15, i32* %18
  br label %19
; <label>:19                                      ; preds = %14, %11, %8, %5, %0
  ret void
}
\end{ccode}

>>>>

\author{David Déharbe, Valério Medeiros Jr.}
\title{Specification of Advanced Types Translation}

@@

The supported advanced types are array and structures. The nodes Array and 
Structure can be used to declare a variable, or  declare an element, or to 
select an element in the array or the structure.

@@

    {\scriptsize

@@

        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Vari}}\\
	\multicolumn{3}{c}{\B{Record}\alt\B{Array}\alt\B{ValueItem} } \\
        \multicolumn{3}{|c|}{Value item : \B{ValueItem}} \\
	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\

@@

@@

\begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data struct : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
        \hline
	\B{base} & : &  \B{Struct} \\
	\B{element} & : & \B{Name}\\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
        \hline
	\B{elements} & : & seq \B{Name}\\
	\B{values} & : & seq \B{Expr}\\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The rule lvalue is:
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an ArrayItem} \THEN  \trad{n}{ArrayItem} \\
& \quad \IF \mbox{\B{n} is a StructItem} \THEN  \trad{n}{StructItem} \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

The expressions have elements represented by pointers that need a special 
care. The attributions between these elements are made  directly by the memory 
address.  However, the expression resolution are made by parts, where the 
partial results are stored to temporary variables. Therefore, the expression 
resolution with pointers needs an additional instruction $LOAD$ for store the 
partial results. This translation is represented by $ValueItem$.

\begin{align*}
\begin{split}
  \trad{n}{\B{ValueItem}} \isdef \\
   & \LET \ell = \lbl  \IN\\
   & \IF \B{n} \mbox{ is \B{StructItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{StructItem}} \\
   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
   & \sep  \ell  \sep t'\\
\end{split}
\end{align*}

@@

Two approaches are possible to support arrays and structures. An approach is 
importing reusable libraries of AtelierB that are reported in the official 
document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
This approach is supported, but the generated code is more complex and it needs 
more intermediate steps to manipulate the data of imported components. 
Moreover, this approach allows dynamic allocation for arrays, but it is 
strongly not recommended for safety-critical systems. A second approach enjoys 
the direct correspondence between B0 native types and LLVM and this approach 
is presented here.

\subsection{Structure} \VGM{It is not implemented yet, because it requires generating an XML example.}
The structure declaration defines a list of variables to be placed under one 
name, allowing several variables to be accessed by one base name. This example  
initializes a structure, and gets and sets one value of xx'balance.

\begin{ccode}
xx := rec(id:1, balance:10);
xx'balance := xx'balance -1;
\end{ccode}
The lines 1 and 2 are represented in translation in LLVM code:

\begin{llvmcode}
%1 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 0
store i32 1, i32 * %1
%2 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
store i32 10, i32 * %2

%3 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
%4 = load i32* %3
%5 = sub nsw i32 %4, 1
%6 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
store i32 %5, i32* %6
\end{llvmcode}

The $StructItem$ rule gets an element value from the struct. The same rule is used on the left side of attribution as a $LValue$ and right side of attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.

\begin{align*}
\begin{split}
  \trad{n}{\B{StructItem}} \isdef \\
   & \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   & \llvm{"\PH{\ell} = getelementptr \%\PH{t}* @\PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
%   & \llvm{\PH{\ell2}  = load  \PH{\SBType(\B{n})}* \PH{\ell1} \nl"} \\
   & \sep  \ell  \sep \SBType(\B{n})\\
\end{split}
\end{align*}

The attribution from record structure to a structure variable, line 1 from last 
B code example, is translated by a sequence of application of rule $StructItem$.

\section{Arrays}

The arrays store a fixed-size sequential collection of elements with the same type. 
They consist of contiguous memory locations, where the lowest address (0) 
corresponds to the first element and the highest address to the last element.
In B, a concrete array is a total function where the domain indexes the range 
elements. Declaration examples of arrays are:
\begin{itemize}
	\item $array1 \in (0..10) \to INT $	(a simple array)
	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
	\item $\{1\vdash1\vdash4, 1\vdash2\vdash5\}  \in (INT \times INT) \times INT$ (an initialised array)
\end{itemize}

\subsection{Relation arrays B0 and LLVM}

@@

This section shows the relation between arrays B0 and LLVM. The sequence of
elements are defining an array, getting an element and setting an
element.

@@

The essential elements of the array are its name, size and type. 
The size of ``e'' array with the indexes (i1,...,in) are defined according 
to the formula:\\
 $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - e.in\_start+1)$.

@@

\VGM{It is not implemented yet, because it requires generating an XML example.}
The following example initialises an new array in C:

@@

The syntax rule\footnote{The underlined element type can be expanded in others types definitions.} used to initialize an array in LLVM is:\\
$@<var\_name> = constant [<size> x \underline{<type>}] [<type> <value>]$.\\
The translation of the last example is:

@@

The LLVM instruction \textit{getelementptr} is used to select an element in
the array or structure. A simple example is:

@@

The following blocks illustrate a C code setting a value to array
and its translation to LLVM.

@@

This example is translated in four steps. First, the variables \%x, \%y, \%z 
are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
Third, the variables are loaded to local variables and converted to 64 bits. 
Finally, the address of array is calculated in local variable \%7 and the value 
7 is stored in it.  
\begin{figure}[h]
\begin{llvmcode}
%x = alloca i32
%y = alloca i32
%z = alloca i32
store i32 2, i32* %x
store i32 3, i32* %y
store i32 4, i32* %z
%1 = load i32* %z
%2 = sext i32 %1 to i64
%3 = load i32* %y
%4 = sext i32 %3 to i64
%5 = load i32* %x
%6 = sext i32 %5 to i64
%7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
store i32 7, i32* %7         ;setting an element
\end{llvmcode}
\end{figure}

@@

\subsection{A  simple example and its LLVM}

@@

define void @Array__set(i32 %ix, i32 %tt) nounwind ssp {
  %1 = alloca i32
  %2 = alloca i32
  store i32 %ix, i32* %1
  store i32 %tt, i32* %2
  %3 = load i32* %1
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %19

; <label>:5                                       ; preds = %0
  %6 = load i32* %1
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %19

; <label>:8                                       ; preds = %5
  %9 = load i32* %2
  %10 = icmp sge i32 %9, 0
  br i1 %10, label %11, label %19

; <label>:11                                      ; preds = %8
  %12 = load i32* %2
  %13 = icmp sle i32 %12, 1000
  br i1 %13, label %14, label %19

; <label>:14                                      ; preds = %11
  %15 = load i32* %2
  %16 = load i32* %1
  %17 = sext i32 %16 to i64
  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
  store i32 %15, i32* %18
  br label %19
; <label>:19                                      ; preds = %14, %11, %8, %5, %0
  ret void
}
\end{ccode}
